Composition-climate paper

### 2016-06-17 Composition_Ecology.Rmd
Broad text edits to reflect changes by @IceAgeEcologist and mine as well.  Also reformatting for Ecol. Letters.

---
title: "Composition-climate paper"
author: "Simon Goring *et al*."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document:
    fig_height: 6
    fig_width: 6
  pdf_document:
    pandoc_args: -V geometry:vmargin=1in -V geometry:hmargin=1in
  md_document:
    variant: markdown_github
csl: ecology.csl
pandoc_args: -H margins.sty
bibliography: goringetal_references.bib
---

Effect of historic land use and climate change on tree-climate relationships for key forest genera in the northern United States
========================================================

@@

library("pander")
library("plyr")
panderOptions('table.style', 'rmarkdown')
```
-->

Abstract
====
* Ecosystems are highly dynamic and contemporary forest have a heavy imprint of land use, while climate changes over decadal and centennial scales.
* Given long time-scale climate change and continued land use conversion, are species-climate relationships today representative of those in past.
* We assess the stability of species-climate relationships using historic PLS data and FIA data.
* We show that species-climate relationships have significantly shifted for many taxa.
* Contemporary trees have shifted to cooler maximum temperatures, warmer winter temperatures, reduced seasonal temperature range, and wetter conditions.
* Relative importance of land use and climate change varies by taxon and climate variable.  
* Shifts in warmer temperature mainly due to land use change, etc.
* Implications sentence or two at the end.

*Historic land use and forest change over the centuries since Euroamerican settlement have significantly changed the structure and composition of forests in the upper Midwestern United States.  The extent to which this change has affected the realized climate niche for key taxa is less well known.*

*Predictions for future species range shifts are largely predicated on correlative models that relate modern species distributions to climate parameters in the modern era.  Given the extent of regional forest change, and recent losses to key forest taxa including Hemlock, Elm and Chestnut, it is possible that certain taxa have shifted their climatic distributions over the last century.  The additional pressure of climatically biased land use conversion for agriculture along the western and southern margins of the upper Midwest means that the shift in climate space may be most pressing in a region of climate space that is critically important for understanding future species responses to climate change.*

*Using contemporary and historical species distributions and climate data we show that modern forest tree genera occupy realized niches with cooler maximum temperatures, higher minimum temperatures and higher rainfall than they have in the past.*

@@

Introduction
==============

Over the past 200 years centennial scale climate variability and land use change have interacted to transform the forests of the northeastern United States. There are clear indications that modern forests, as represented by the Forest and Inventory Analysis show greater homogeneity and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization].

This shift has occured over a period during which climate has begun to show the effects of anthropogenic warming, and a period in which we have increasingly used the relationship between modern tree species distributions and climate to understand future responses of tree species and forest assemblages to climate change [@iverson2013tree;@iverson1998predicting].  One of the most significant changes to forest tree distributions in the upper midwest is the loss of extensive cover through land use change [@rhemtulla2009legacies], particularly along the historical prairie and savanna margins of the region (Figure 1f), where extensive land use conversion has left little natural land.

@@

land_use_plot <- natural[[2]] + 
  geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
  coord_equal(xlim = ext[1:2], ylim=ext[3:4]) + 
  scale_fill_manual(values = c('#ff7f00', '#e41a1c', '#33a02c')) +
  theme(legend.position = 'none')

source('R/figures/climate_ts.R')

@@

ggsave(plot = arrangeGrob(climate_values$plot, land_use_plot, nrow=1, widths = c(3, 4)), filename='figures/figure_one.tiff')

```
![figone](figures/figure_one.tiff)

**Figure 1.** *(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.  Low temperatures in the 1970s result in lower T~max~ during the modern era, but climate change has resulted in higher T~min~ values than the early-century normals, along with increasing P~ann~.  PDSI shows strong coherence with P~ann~ in this region. (f) Patterns of land use change in the upper Midwestern United States. Data from the NLCD [@jin2013comprehensive].*

The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].  Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.  By deliniating three types of interaction: (1) compounding, (2) confounding and (3) contradicting, it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to broader regional change.

Compounding effects occur when both climatic change and land use change are biased in the same direction, *e.g.*, warming climate is compounded by land use change that is biased towards warmer environments.  In the case of warming, a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.  This pattern is seen for butterfly species in the northern Sierra Nevada mountains of California, where warming is compounded by land use conversion at the lowest elevations [@forister2010compounded].  Confounding effects may cause little overall effect on species distributions since land use pressure falls within the central distribution for the environmental gradient. Confounding effects may increase the variability of predictive models, but do little to the overall envelope.  Garry oak (*Quercus garryana*) distributions in the Pacific Northwest may suffer from confounding effects.  The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining;@pellatt2014environmental], thus predictive models [@pellatt2012using] may show greater uncertainty than if complete historical distributions had been used. Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in one tail of the climate distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the other end of the tail excludes the taxa from those regions.  Here we might think of downward migrations of some bird species in the Sierra Nevadas in response to shifts in precipitation [@tingley2012push], where downward migration may be limited by land use conversion in the lowlands.

In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al. [-@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical]. Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies] we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.  Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

Interaction between environmental gradients, land use patterns and taxa are likely to vary by region and environmental variable of interest. Species in the north may experience a confounding land use effect with respect to temperature, northward shifts due to warming temperatures may not have significantly manifested themselves, and land use conversion is likely to be heterogenous with respect to temperature since there is little agriculture in northern Wisconsin and northeastern Minnesota.  Thus, with respect to temperature we expect to see no shift in distribution along temperature gradients.

With respect to precipitation, particularly toward the forest-prairie margin, where land use conversion continues [@wright2013recent] we would expect to see compounding effects of land use and climate change. Northern species should be shifting away from drier habitat while land use in the northern prairie-forest margin intensifies, shifting species further than expected along the precipitation gradient.

For southern tree species, land use conversion is primarily, and has primarily, occured along the southern margins of the region, parallel to temperature and precipitation gradients. This should result in compounding land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the loss of suitability and land use conversion along the southern margin of the species' range.

Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting and confounding land use effects may result in the impression of narrower climatic tolerances, or greater predictive uncertainty. Tree species are responding to climate change within forested landscapes, but the response is inconsistent. Differences in the distributions of mature trees and seedlings show responses in western North America [@monleon2015evidence], but apar to indicate range contraction, rather than northward expansion in eastern temperate North America [@zhu2012failure]. Modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts, potentially clouding our interpretation of these patterns. For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, dut to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.

We use gridded climate data products (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2015preset] to develop climate-vegetation relationships for 15 major tree taxa in the upper Midwestern United States. By examining the interactions between climate, species distributions and land use change since EuroAmerican settlement we examine the possible implications of range shifts for future estimates of species resillience to climate change across the region.

Methods
===============
We pair pre-settlement era vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.  The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.  At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.  Goring et al. [-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.  When examining relationships between FIA and PLS data, Goring et al. [-@goring2015preset] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.

Another complication is the structure of the FIA dataset.  While the PLS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].  As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA. This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations, agency that ultimately leads urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological].

The analysis presented here follows a blocked sampling method. We generate probability density functions along historical climate gradients for taxa within the PLSS and for the same PLSS taxa (and point locations) using climate data for the modern era. We do the same for FIA presence (climate at the PLSS and modern eras), and for anthropogenic land use classes. Climate data is derived from monthly PRISM data from 1895 - 2014. A 30-year climate normal was generated for modern July temperature ($T_{jul}$), January temperature ($T_{jan}$), the continentality (difference between $T_{jul}$ and $T_{jan}$, $T_{diff}$), and annual precipitation ($P_{ann}$).

The choice of a thirty year normal for the PLS data is potentially problematic.  The time transgressive nature of the PLS Survey means that some of the PLS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset;@white1983history], and for much of this time period no climatological records exist. Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period. As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice.  Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions tell us that there has been a trend of increasing temperatures (approximately 0.3^o^C in the Northern Hemisphere) since the 1800s, with pollen-based reconstructions showing even higher rates of change in the Midwestern United States [@viau2006millennial], with tree rings showing cooling at the 1800s to warming from the 1900s [@ahmed2013continental].

By generating kernel density estimates of taxon presence along climate gradients drawn from each of the four states of climate and vegetation ($PLSS_{veg}$-$PLSS_{clim}$, $PLSS_{veg}$-$FIA_{clim}$, $FIA_{veg}$-$PLSS_{clim}$, $FIA_{veg}$-$FIA_{clim}$) we can use Hellinger distance to estimate the effects of changing climate ($d_{H}$), or the effect of changing land use (d_H_ between kernel densities in the same row of Figure 2).  In particular, the effect of changing land use and climate (d_H_ between the top left and bottom right in Figure 2) can be parsed as the sum of differences in the left hand column (testing climate change) and the top row (testing land use only).  The relative magnitudes of these differences gives us an indication of the influence of either land use or climate in driving shifts in species distributions along the climate gradients in the Upper Midwest.

Results
---------------

Mean temperatures in the region show an average increase of `r round(clim_change[3], 2)`^o^C since the 1895-1925 normal.  Annual precipitation has increased slightly (`r round(clim_change[1], 0)`mm), and maximum annual temeprature has declined slightly (`r round(clim_change[2], 2)`^o^C).  The most significant change has been in annual minimum temperatures, which have increased by `r round(clim_change[4], 1)`^o^C since the 1895-1925 normal.  This broad scale increase in winter temperatures is widely accepted, and results in an overall shift in the winter 'climate space' for the region.

The Public Lands data is spatially extensive, sampling occured across the region in a regular pattern, while the FIA is limited to regions with forest cover.  As such, the extent of points is not overlapping. Given the extensive use of FIA data in generating and estimating climate responses of tree taxa and forest types to changing climate, the use of the FIA data as an estimate for tree species distributions is acceptable fo estimating shifts in climate space, and attributing the loss or gain of climate niche space due to land use and climate variability and change during the 20^th^ century.

@@

loss_table[,1:3] <- round(tran(loss_table[,1:3], method = 'proportion') * 100,1)

ggsave(plot = loss_melt[[2]], filename = 'figures/Figure_panelloss.tiff',
       dpi=300, width=6, height = 3)
```

![figuretwo](figures/Figure_panelloss.tiff)
**Figure 2.**.  *Gain loss and continuous presence for PLSS and FIA tree taxa. Loss is assessed as cells that contained presence during the PLSS era, and no presence in the FIA data (light blue). Continuous presence is indicated by dark gray, where both FIA and PLSS cells indicate presence.  Few taxa show novel presence (red), although taxa such as poplar, ash and basswood show gains of over 10%.*

**Table 1.** *Gain and loss of the various forest tree types of interest since the pre-settlement era. The first three columns represent percentages of the total area occupied by the taxon since settlement. Modern and Historical values represent 100s of km^2^ occupied by the tree taxon within either the FIA or PLSS data.*

```{r, results='as.is', echo=FALSE}
pandoc.table(loss_table, justify = "left", style = 'simple')
```

```{r densities, warning=FALSE, message=FALSE, echo=FALSE, dpi=150}
source('R/resampling_climate.R')
vegclim_table <- resample_climate(climate, agg_dens, pls_data)

source('R/figures/climate_boxplots.R')
boxplot_clim <- clim_boxplots(clim_table = vegclim_table)

>>>>

---
output:
  word_document: null
  pdf_document:
    pandoc_args: -V geometry:vmargin=1in -V geometry:hmargin=1in
  md_document:
    variant: markdown_github
csl: ecology.csl
pandoc_args: -H margins.sty
bibliography: goringetal_references.bib
---

@@

library("pander")
library("plyr")

panderOptions('missing', '--')
panderOptions('digits', 4)
panderOptions('keep.trailing.zeros', TRUE)
panderOptions('table.continues', NULL)
pander::panderOptions('table.style', 'rmarkdown')
pander::panderOptions('table.split.table', 300)
grDevices::ps.options(fonts = 'serif')

```
-->

# Effect of historic land-use and climate change on tree-climate relationships in the northern United States

**Statement of Authorship**

* Simon J Goring: Conceived paper, performed analysis, developed figures and wrote the paper
* John "Jack" W Williams: Wrote the paper

**Data Accessibility**

* All data for this manuscript, along with code and version-control is available at http://github.com/SimonGoring/CompositionClimate.  In addition, an archived copy of the published version is available with the following DOI.

**Authors**

* Simon J Goring
  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
* Jack W Williams
  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706
  * Center for Climatic Research, University of Wisconsin, Madison, 1225 W Dayton St, Madison WI 53706

**Keywords**

* climate change, land use, historical ecology, realized niche, anthropocene, public land survey, forest inventory and analysis

**Running Title**

* Land use change forces climate shifts in trees

**Article Type**

* Letters

**Number of Words**

* Main Text:
* Text boxes:

**Number of References**

**Number of Figures**

**Corresponding Author**

* Simon J Goring
  * Department of Geography, University of Wisconsin, Madison, 550 N Park St, Madison WI 53706

# Abstract

Forest inventory data are essential for understanding tree species distributions along environmental gradients, and for projecting forest responses to climate change.  However, the stability and representativeness of contemporary tree-climate distributions is poorly understood.  Here we show that historic land use and climate change over the last two centuries have significantly altered tree-climate relationships, using contemporary and historical data for climates and 15 common tree genera in the northern US. Realized niches have shifted towards cooler maximum temperatures, higher minimum temperatures, and higher rainfall. A new attribution method indicates that both historic climate change and land use produced these shifts, with the relative importance varying among genera and climate variables.  The most common climate/land-use interaction is compounding, in which historic land-use exacerbates climatic shifts toward warmer or wetter distributions. Compounding interactions are critical because they suggest that contemporary-based models of species distributions may underestimate species resilience to climate change.

/pagebreak

@@

# Introduction

*The introduction should summarize briefly the background and aims, and end with a very brief statement of what has been achieved by the work.*

Global change ecologists and biogeographers use contemporary distributions of species, and, increasingly, paleohistorical datasets to map distributions of species and species traits within environmental space REFS. These distributions of species within climate space, combined with future climate scenarios and species distribution models, can be used to assess species exposure and sensitivity to climate change, identify species and habitats of concern, set conservation priorities, and prepare for shifts in habitat suitability caused by climate change  [@pellatt2012using;@iverson2013tree;@iverson1998predicting;@anderson2013nichemodels;@guisan2013predicting]. Forest inventory datasets, such as the US Forest Inventory Analysis (FIA), are particularly rich observational datasets that have, *e.g.* been used to assess early signals of range shifts via the distributions of mature trees and seedlings [@monleon2015evidence;zhu2012failure] and to parameterize models of current and future tree distributions [@iverson2008estimating;@iverson1998predicting].

Increasingly, ecological models and their predictive capacity are being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].  Future projections that rely only on contemporary observational data are limited by a narrow time window that may obscure greater plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches that lie outside the current realized environmental space [@jackson2000responses], miss the responses of species to past no-analog climates [@veloz2012no], and fail to detect the effects of historical legacies and contingency on species distributions [@svenning2015influence]. Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence]. Habitat loss through human agency can impact apparent species distributions at timescales of thousands [@sarmento2013impacts] to hundreds of thousands of years [@ivory2016niche]. In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.  In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close].

Over the last several centuries, climate change and land use have both substantially affected species distributions and vegetation structure [@gehrig2007tree] REFS. Understanding the interacting effects of land-use change and climatic change on species-climate relationships is of particular importance, because anthropogenic land use is highly structured and differentially occupies particular habitat types and portions of climate space.  Hence, species distributions and protected natural areas often represent a biased selection from the available environmental space [@scott2001nature].  This biasing can confound studies that use contemporary distributional data to predict species responses to future climate scenarios [@pyke2004habitat].

The interactions between climate and land-use change can be classified into three types: (1) compounding, (2) confounding and (3) counteracting [modified from @pyke2004habitat]. Compounding effects occur when climatic and land-use change cause species distributions to shift in the same direction.  For example, for butterfly species in the northern Sierra Nevada Mountains of California, warming has been compounded by land-use conversion at the lowest elevations, both driving species upslope [@forister2010compounded]. Confounding effects may cause little overall change on species distributions, with land-use pressure falling within the central distribution for the environmental gradient.  The overall spatial and climatic distribution of Garry oak (*Quercus garryana*) in the Pacific Northwest has changed little over the past several centuries, but land-use conversion has excluded it from the deep-soil sites it formerly occupied [@macdougall2004defining;@pellatt2014environmental].  Hence, predictive models for Garry oak [@pellatt2012using] using only contemporary data may be confounded by this range-wide interaction between human land use and soil characteristics. Counteracting effects occur when climate and land use change act in opposite directions, narrowing the available geographic and climatic space available for species. For example, bird species in the Sierra Nevadas should show downslope migration in response to shifts in precipitation [@tingley2012push], but this downslope migration is limited by land-use conversion in the lowlands.

@@

land_use_plot <- natural[[2]] + 
  geom_tile(data = land_use, aes(x = x, y = y, fill = class)) +
  coord_equal(xlim = ext[1:2], 
              ylim = ext[3:4]) + 
  scale_fill_manual(values = c("#ff7f00", "#e41a1c", "#33a02c")) +
  theme(legend.position = "none")

source("R/figures/climate_ts.R")

@@

ggsave(plot = arrangeGrob(climate_values$plot, land_use_plot, nrow = 1, widths = c(3, 4)), 
       filename = "figures/figure_one.tiff")

```
![figone](figures/Fig_1_Composite.png)

`r fig_nums("land_use_clim", "*(a-e) Climate change in the Upper Midwest over the last two centuries using PRISM data and North American Drought Atlas PDSI reconstructions.  Low temperatures in the 19XXs result in lower T~max~ during the modern era, but climate change has resulted in higher T~min~ values than the early-century normals, along with increasing P~ann~.  PDSI shows strong coherence with P~ann~ in this region. (f) Patterns of land-use change in the upper Midwestern United States. Land use data from the NLCD [@jin2013comprehensive].*")`

In the upper Midwestern US, tree distributions and community composition over the last several centuries have been heavily influenced both by the land uses associated with EuroAmerican Settlement in the 1800s [@goring2015preset;@schulte2007homogenization;hanberry2012comparison;@rhemtulla2009historical] and by the climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response;@paquette2013climatic] and current anthropogenic warming [@wicci2011climate].  Modern forests, as represented by the Forest and Inventory Analysis [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System, a survey that covered the western United States prior to widespread land-use conversion in the late 1800s [@goring2015preset;@schulte2007homogenization]. There has been widespread conversion of wildland ecosystems to agricultural and urban land use [@rhemtulla2009historical], particularly along the historical prairie and savanna margins of the region (`r fig_nums("land_use_clim", display = "cite")`f).  Forested regions in the northern Great Lakes States have been also affected by widespread timbering during the late 19th and early 20th-century and ongoing land management REF.  In northern forested regions, we may expect to see expanded ranges of early successional species and reduced abundances of late-successional species, following widespread logging.  Additionally, some tree species, *Ulmus americana* and *Castanea dentata* have experienced rangewide population collapses caused by the introduction of exotic pathogens REFS. These land use changes are superimposed on strong regional gradients of temperature and precipitation in the upper Midwest and two major ecotones, one from prairie/savanna to closed forest, in an approximately west to east direction, and one from conifer dominated, sub-boreal forests to deciduous forest in a north to south direction.

Here we show that the climatic distributions of tree genera in the upper Midwest have significantly shifted over the last 100 years as a result of historic climate change, land use, and pathogen outbreaks, and assess the relative impacts of regional climate shifts and vegetation change in driving these changes.  To this end, we use gridded historic and contemporary climate observations (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2015preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates and develop a new method, based on Hellinger distances among tree distributions, to quantify the total change and relative importance of historic vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of land-use conversion on species resilience to climate change as predicted by species-distribution models generated from modern distributional data.

# Materials and Methods

*This section should contain sufficient detail so that all procedures can be repeated (in conjunction with cited references). A checklist is provided so that authors can check that their methods report details which our editors regard as essential. This checklist is available here. Where specific equipment and materials are named, the manufacturer's name, city and country should be given (generally in parentheses after first mention).*

## Historical and Modern Vegetation Data:

`r tab_nums("genera_table", "Tree genera used in the study, along with total spatial coverage (km2)in the pre-settlement and modern eras based on presence within an 8x8km grid cell.  Genera are grouped into broadleaved and needleleaved taxa, then ordered alphabetically.  The total regional coverage is 427,000 km^2^ for the PLS data and 331,000 km^2^ for the FIA data.")`

```{r, echo = FALSE, message = FALSE, warning=FALSE, results='asis'}

area_table <- data.frame(Genera = colnames(pls_data)[-1],
                         `PLS Coverage\nkm^2^` = prettyNum(colSums(pls_data > 0) * 64, 
                                                    big.mark = ',', digits = 3)[-1],
                         `FIA Coverage\nkm^2^` = prettyNum(colSums(agg_dens > 0) * 64, 
                                                    big.mark = ',', digits = 3)[-1])

# Capitalize the taxa and sort them by PFT:
area_table[,1] <- stringi::stri_trans_totitle(area_table[,1])
area_table <- area_table[c(9,6,1,15,8,2,13,11,14,
                           5,4,3,10,7,12),]

rownames(area_table) <- NULL

pander::pandoc.table(area_table, style = 'rmarkdown')

Settlement-era forest data are based on survey data from the Public Land Survey System [PLSS: @white1983history;@schulte2001original;@goring2015preset], aggregated to a 64 km^2 grid across the upper Midwestern United States [@goring2015preset]. We briefly review the methods here. PLSS data originally were collected on a 1 x 1 mi grid across the Upper Midwest from 1830 - 1904 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey, point surveyors noted the closest two to four trees and recorded their common name, distances to the survey point, diameters at breast height and azimuth to the tree. Goring et al. [-@goring2015preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2015preset]. Using the count data produced by Goring et al. [-@goring2015preset], we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera (Table `r tab_nums("genera_table", display = "cite")`). Common names for certain taxa, such as "Ironwood" can not be clearly resolved to a single tree genera, as such we combine *Ostrya* and *Carpinus* (ironwood), and *Thuja* and *Juniperus* (cedar). Here presence is defined as at least one recorded tree in any 8x8 km grid cell.

Modern forest data are from the US Forest Inventory and Analysis (FIA) program, which began in the 1930s and provide detailed inventories of contemporary tree distributions [@iverson2013tree;@zhu2012failure]. FIA datasets are publicly available surveys of forest composition (for all regions with > 10% forest cover) [@woudenberg2010forest].  FIA uses a nationally standardized sampling procedure, for all land with greater than 10% forest cover.  Each Phase 2/3 FIA plot  consists of four, 7.32m radius, circular sub-plots which are sampled intensively [@woudenberg2010forest].

Comparisons between historical and modern vegetation must address differences in sampling design between the PLSS and FIA datasets [@goring2015preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here the primary tool for maximizing comparability is via aggregation, both by spatially aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  For example,  the FIA contains fewer plots per 8 km cell, with more trees per plot than the PLSS data. This leads to higher cell-to-cell variability in the FIA data, but these are not expected to affect the regional-scale analyses here, particularly because we are investigating presence/absence patterns. Additionally, FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing the spatial extent of many tree taxa to be lower in the FIA than in the PLSS datasets (`r tab_nums("genera_table", display="cite")`). Hence, scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However the practice of using FIA data as a surrogate for tree presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], so this complication is a general one faced by forest ecologists and global change ecologists.  In prior comparative analyses of the PLSS and FIA data, Goring et al. [-@goring2015preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect.

## Historical and Modern Climate Data

Historical and modern climate data are derived from monthly PRISM data for 1895 to 2014 CE [@prism]. The PRISM climate variables are available at an 800m resolution and have been interpolatedfrom climate stationlocations using elevation, aspectand other data.  Estimates for July temperature ($T_{jul}$), January temperature ($T_{jan}$), the annual temperature range ($T_{diff}$), and annual precipitation ($P_{ann}$) were obtained from PRISM and resampled to the 8 x 8km grid used for PLSS and FIA data using bilinear interpolation. Two periods were selected for the calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.  There is an unavoidable temporal mismatch between the historical climate data and the PLSS survey period, on the order of 50 to 70 years.  Few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period.  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we check the instrumental record against tree-ring-based paleoclimate reconstructions from the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).

## Shifts in Species Distributions in Climate Space

![modelgrid](figures/figureFour_panels.tiff)

`r fig_nums("fourPanel", "(a) Diagram of the 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on the observed shifts in tree-climate relationships. We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for Larix for each of the four combinations of climate and vegetation states, plotted relative to Tmax (b), Tdif (c), Tmin (d), and Pann (e).")`

Distributions of tree genera presences along climate gradients are described as probability density functions for both the pre-settlement and modern eras.  Differences in distributions are quantified using Hellinger distances [@matthes2015benchmarking] to compare the probability density functions among the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate and the subscripts M and H represent modern and historical conditions, respectively). The total shift ($d_{tot}$) between presettlement conditions and modern conditions is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ densities (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The effect of historical climate change on shifting climate distributios of tree genera is determined by calculating the Hellinger distance ($d_{C}$) is determined by calculating the Hellinger distance between PLSS and FIA tree distributions when both are overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{M}C_{H}$; `r fig_nums("fourPanel", display = "cite")`a; left column). We develop an attribution index, ($\delta_{d}=d_{c} - d_{l}$), such that negative values of$\delta_{d}$ indicate stronger influence of historical land use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.

To assess whether the effects of historical land use and climatic changes on a genera's climatic distribution is compounding, counteracting or confounding we perform a set of t-tests, using the conservative Bonferroni correction to account for multiple tests.  For each taxon and climate variable pair we test whether $V_{H}C_{H}$ and $V_{H}C_{M}$ (climate change), and $V_{H}C_{H}$ and $V_{M}C_{H}$ (land-use change) distributions are significantly different.   If both changes are significant and of the same sign, then the shift is defined as a compounding change.  If both are significant and are in opposite directions then the change is defined as counteracting.  In the case that either or both changes are not significant then the change is defined as confounding.

Analysis in this paper is undertaken using R and R Markdown.  All code and raw data is provided in supplemental material and hosted on GitHub (http://github.com/SimonGoring/Composition_Climate).  Analysis makes use of the R packages colorspace, captioner, raster, ggplot2, mgcv, reshape2, plyr, gridExtra, rgdal, RColorBrewer, analogue, and maptools.  Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape-level features.

# Results

*The Results section should present the experiments that support the conclusions to be drawn later in the Discussion. The Results section should conform to a high standard of rigour. Extended lines of inference, arguments or speculations should not be placed in the Results.*

Regionally, January (minimum) temperatures have increased by `r round(clim_change[4], 1)`^o^C since the 1895-1925 normal `r fig_nums("land_use_clim", display = "cite")`, maximum annual temperature has declined (`r round(clim_change[2], 1)`^o^C), while mean annual temperatures increased by `r round(clim_change[3], 1)`^o^C (not shown). Because of the large shift in minimum temperatures, Tdiff has changed by .  Annual precipitation in the region has increased slightly (`r round(clim_change[1], 1)`mm).  The Palmer Drought Severity Index is closely coherent with Pann trends from 1895 onward and indicate no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^ century climates are a reasonable proxy for mid 19^th^ century climates.

@@

loss_table[,1:3] <- round(tran(loss_table[,1:3], method = "proportion") * 100,1)

ggsave(plot = loss_melt[[2]], filename = "figures/Figure_panelloss.tiff",
       dpi = 300, 
       width = 6, 
       height = 3)
```

![figuretwo](figures/Figure_panelloss.tiff)

`r fig_nums("gain_loss", "Maps showing areas of gain, loss, and continuous presence of tree genera. 'Loss' indicates 8x8km grid cells where PLSS data indicated presence of a genus but FIA data do not record presence (light blue). 'Gain' indicates areas where a genus is absent in the PLSS data but present in the FIA data (red). 'Continuous presence' represents locations where both FIA and PLSS data indicate presence (dark gray). Losses are more common than gains, particularly in the southwestern portion of tree ranges.  Few taxa show gains, although Populus, Fraxinus and Tilia show gains of over 10% (Table 2).")`

```{r echo = FALSE}
tab_nums("tab_gain_loss", "Gain and loss of the analyzed tree genera since the pre-settlement era. The first three columns are percentages relative to the total area occupied by the taxon in the presettlement and modern eras. Modern and Historical indicate represent range area (100s of km2) occupied by the tree taxon for the pre-settlement or modern era.  Taxa are sorted by the percent area lost (Column 3).", display = FALSE)
```
